---
title: "XGBoost Project for ADM"
author: "Matthias Quinn"
date: "11/08/2021"
output: 
  html_document: 
    toc: yes
    theme: cerulean
    fig_caption: yes
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)
```

# Exploring and Applying XGBoost

## Abstract:

This project will focus on the famous XGBoost system and its application on moderate and large datasets. Starting with a history of the system, then exploring the algorithm itself, and finally ending with two applications, this project will hopefully provide a framework on which to base further research and applications.

## History:

From Tianqi Chen and Carlos Guestrin in March of 2014, XGBoost has been a monumental system in the field of advanced analytics and machine learning. XGBoost was initially a research project started by Tianqi Chen - and later Carlos Guestrin - at the University of Washington, who presented their paper at SIGKDD Conference in 2016. Since then, version 1.5.0 is available for public usage and has been implemented in more than 5 languages including: C++, Python, R, Java, Scala, and Julia. In addition, the system is available to all modern operating systems, including: Windows, OS X, Linux, and a variety of cloud platforms.

## The XGBoost Algorithm:

## Application 1 - Adult Income:

## Application 2 - Forest Cover Types:

Given forestry data from four wilderness areas in Roosevelt National Forest, classify the patches into one of $7$ cover types, listed below:

1.  Aspen
2.  Cottonwood/Willow
3.  Douglas-fir
4.  Krummholz
5.  Lodgepole Pine
6.  Ponderosa Pine
7.  Spruce/Fir

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(caret)
library(nnet)
library(xgboost)
library(mlbench)
```

Reading in the forest cover-type dataset.

```{r}
X <- read_excel("C:/Users/miqui/OneDrive/Datasets/covtype.xlsx", 
    sheet = "X")
y <- read_excel("C:/Users/miqui/OneDrive/Datasets/covtype.xlsx", 
    sheet = "y")
```

Combine the above two data frames into $1$ for easier access:

```{r}
df <- bind_cols(X, y)
```

Turn the response variable `Cover_Type` into a factor with $7$ levels.

```{r}
df$Cover_Type <- factor(df$Cover_Type)
```

## Exploratory Data Analysis:

+------------------------------------+--------------+---------------------------------------------------------+
| Variable                           | Type         | Description                                             |
+====================================+==============+=========================================================+
| Elevation                          | Numeric      | Elevation in meters                                     |
+------------------------------------+--------------+---------------------------------------------------------+
| Aspect                             | Numeric      | Aspect in degrees azimuth                               |
+------------------------------------+--------------+---------------------------------------------------------+
| Slope                              | Numeric      | Slope in degrees                                        |
+------------------------------------+--------------+---------------------------------------------------------+
| Horizontal_Distance_To_Hydrology   | Numeric      | Horizontal distance to nearest surface water features   |
+------------------------------------+--------------+---------------------------------------------------------+
| Vertical_Distance_To_Hydrology     | Numeric      | Vertical distance to nearest surface water features     |
+------------------------------------+--------------+---------------------------------------------------------+
| Horizontal_Distance_To_Roadways    | Numeric      | Horizontal distance to the nearest roadway              |
+------------------------------------+--------------+---------------------------------------------------------+
| Hillshade_9am                      | Numeric      | Hillshade index at 9 a.m.                               |
+------------------------------------+--------------+---------------------------------------------------------+
| Hillshade_Noon                     | Numeric      | Hillshade index at noon                                 |
+------------------------------------+--------------+---------------------------------------------------------+
| Hillshade_3pm                      | Numeric      | Hillshade index at 3 p.m.                               |
+------------------------------------+--------------+---------------------------------------------------------+
| Horizontal_Distance_To_Fire_Points | Numeric      | Horizontal distance to nearest wildfire ignition points |
+------------------------------------+--------------+---------------------------------------------------------+
| Wilderness_Area (4 binaries)       | Categorical  | Wilderness area designation                             |
+------------------------------------+--------------+---------------------------------------------------------+
| Soil_Type (40 binaries)            | Categorical  | Soil type designation                                   |
+------------------------------------+--------------+---------------------------------------------------------+
| Cover_Type (7 types)               | Categorical  | Forest cover type designation                           |
+------------------------------------+--------------+---------------------------------------------------------+

: Variables

## Feature Engineering:

It would be nice to have a name associated with each cover type.

```{r}
table(df$Cover_Type)
```

```{r}
df <- df %>%
  mutate(CoverName =
           case_when(Cover_Type == 1 ~ "Spruce.fir",
                     Cover_Type == 2 ~ "Lodgepole.Pine",
                     Cover_Type == 3 ~ "Ponderosa.Pine",
                     Cover_Type == 4 ~ "Cottonwood.Willow",
                     Cover_Type == 5 ~ "Aspen",
                     Cover_Type == 6 ~ "Douglas.fir",
                     Cover_Type == 7 ~ "Krummholz",
                     TRUE ~ "Other"))
```

Combining the multiple, binary wilderness area type variables into $1$.

```{r}
df <- df %>%
  mutate(WildernessArea =
           case_when(Wilderness_Area_0 == 1 ~ 0,
                     Wilderness_Area_1 == 1 ~ 1,
                     Wilderness_Area_2 == 1 ~ 2,
                     Wilderness_Area_3 == 1 ~ 3,
                     TRUE ~ NaN))
table(df$WildernessArea)
```

Combining the multiple, binary soil-type variables into $1$.

```{r}
df <- df %>%
  mutate(SoilType =
           case_when(Soil_Type_0 == 1 ~ 0, Soil_Type_1 == 1 ~ 1,
                     Soil_Type_2 == 1 ~ 2, Soil_Type_3 == 1 ~ 3,
                     Soil_Type_4 == 1 ~ 4, Soil_Type_5 == 1 ~ 5,
                     Soil_Type_6 == 1 ~ 6, Soil_Type_7 == 1 ~ 7,
                     Soil_Type_8 == 1 ~ 8, Soil_Type_9 == 1 ~ 9,
                     Soil_Type_10 == 1 ~ 10, Soil_Type_11 == 1 ~ 11,
                     Soil_Type_12 == 1 ~ 12, Soil_Type_13 == 1 ~ 13,
                     Soil_Type_14 == 1 ~ 14, Soil_Type_15 == 1 ~ 15,
                     Soil_Type_16 == 1 ~ 16, Soil_Type_17 == 1 ~ 17,
                     Soil_Type_18 == 1 ~ 18, Soil_Type_19 == 1 ~ 19,
                     Soil_Type_20 == 1 ~ 20, Soil_Type_21 == 1 ~ 21,
                     Soil_Type_22 == 1 ~ 22, Soil_Type_23 == 1 ~ 23,
                     Soil_Type_24 == 1 ~ 24, Soil_Type_25 == 1 ~ 25,
                     Soil_Type_26 == 1 ~ 26, Soil_Type_27 == 1 ~ 27,
                     Soil_Type_28 == 1 ~ 28, Soil_Type_29 == 1 ~ 29,
                     Soil_Type_30 == 1 ~ 30, Soil_Type_31 == 1 ~ 31,
                     Soil_Type_32 == 1 ~ 32, Soil_Type_33 == 1 ~ 33,
                     Soil_Type_34 == 1 ~ 34, Soil_Type_35 == 1 ~ 35,
                     Soil_Type_36 == 1 ~ 36, Soil_Type_37 == 1 ~ 37,
                     Soil_Type_38 == 1 ~ 38, Soil_Type_39 == 1 ~ 39,
                     TRUE ~ NaN))
table(df$SoilType)
```

```{r}
ggplot(df, aes(x=SoilType, fill=CoverName)) + geom_bar(aes(y=(..count..)/sum(..count..))) + ggtitle('Proportion of Observations by Soil Type') + ylab('Proportion') + xlab(NULL) + theme(legend.title=element_blank(), legend.position='bottom') +
scale_x_discrete(breaks = 1:40) + theme(text = element_text(size=16))
```

## Modeling:

### Choose the variables that will be fed to the upcoming models. {#choose-the-variables-that-will-be-fed-to-the-upcoming-models.}

```{r}
df <- df %>%
  select(Elevation, Aspect, Slope, Horizontal_Distance_To_Hydrology,
         Vertical_Distance_To_Hydrology, Horizontal_Distance_To_Roadways,
         Hillshade_9am, Hillshade_Noon, Hillshade_3pm, Horizontal_Distance_To_Fire_Points, WildernessArea, SoilType, Cover_Type,
         CoverName)
```

### Split the main data frame into train-test sets: {#split-the-main-data-frame-into-train-test-sets}

```{r}
set.seed(1234)
inTrain <- createDataPartition(df$Cover_Type,
                                  p = .75, 
                                  list = FALSE, 
                                  times = 1)
training <- df[inTrain, ]
testing <- df[-inTrain, ]
```

### Setting up Cross-Validation: {#setting-up-cross-validation}

```{r}
trControl <- trainControl(method = "repeatedcv",
                          repeats = 2,
                          verboseIter = TRUE,
                          allowParallel = TRUE,
                          preProcOptions = c("center", "scale"),
                          classProbs = TRUE) # For ROC calculation
```

### Initial Hyper-parameters for XGBoost: {#initial-hyper-parameters-for-xgboost}

```{r}
xgbHyperParams <- expand.grid(
                        nrounds = 2,
                        max_depth = c(5, 10),
                        eta = 0.05,
                        gamma = 0.01,
                        colsample_bytree = 0.75,
                        min_child_weight = 0,
                        subsample = 0.5)
```

### Fitting a Multinomial Logistic Model:

The following multinomial model will take more than 15 minutes to run, so run at your own risk.

```{r}
set.seed(5678)
multinomMod <- train(CoverName ~ Elevation,
                 data = training,
                 method = "multinom",
                 metric = "Accuracy",
                 trControl = trControl,
                 verbose = FALSE)
```

```{r}
multinomMod
```

### Fitting our XGBoost Model:

```{r}
set.seed(1234)
xgbMod1 <- train(CoverName ~ Elevation,
                 data = training,
                 method = "xgbTree",
                 metric = "Accuracy",
                 trControl = trControl,
                 tuneGrid = xgbHyperParams,
                 verbose = FALSE)
```

### Examining the Results:

```{r}
xgbMod1
```

```{r}
ggplot(xgbMod1)
```

## Model Comparison:

```{r}
resamps <- resamples(list(mlr = test, xgb = xgbMod1))
summary(resamps)
```

```{r}
xyplot(resamps, what = "BlandAltman")
```

## References:

[XGBoost: A Scalable Tree Boosting System](https://arxiv.org/abs/1603.02754)

[UCI: Cover Type Data set](https://archive.ics.uci.edu/ml/datasets/Covertype)
